### 2016-06-30 GPCE2016/sections/CaseStudy.tex
grammar

As the first application for Classless Java, we illustrate a useful
programming pattern that improves modularity and extensibility of
programs. This useful pattern is based on an existing solution to
the \emph{Expression Problem} (EP)~\cite{wadler98expression}, which is a well-known
problem about modular extensibility issues in software evolution. Recently, a
new solution~\cite{eptrivially} using only covariant type refinement was
proposed. When this solution is modeled with interfaces and default methods, it
can even provide independent extensibility~\cite{zenger05independentlyextensible}: the ability to assemble a system
from multiple, independently developed extensions. Unfortunately, the
required instantiation code makes a plain Java solution verbose and cumbersome
to use. The \mixin annotation is enough to remove the boilerplate code, making
the presented approach very appealing. Our last case study, presented
in Section~\ref{subsec:int}, is essentially a (much larger) application of this
pattern to an existing program. Here we illustrate the pattern in the
much smaller Expression Problem.

\paragraph{Initial System}
In the formulation of the EP, there is an initial system that models
arithmetic expressions with only literals and addition, and an initial
operation \texttt{eval} for expression evaluation.
As shown in Figure~\ref{fig:ep}, \texttt{Exp} is the common
super-interface with operation \texttt{eval()}
inside. Sub-interfaces \texttt{Lit} and \texttt{Add} extend interface
\texttt{Exp} with default implementations for the \texttt{eval} operation. The
number field \texttt{x} of a literal is represented as a getter method
\texttt{x()} and expression fields (\texttt{e1} and \texttt{e2}) of an addition
as getter methods \texttt{e1()} and \texttt{e2()}.

@@

\paragraph{Extending the Query Language} The previous query language can be extended with a new feature
\texttt{orderBy} which orders the result records by a field that users
specify. With \mixin programmers just need to extend the interface \texttt{Database} with new
features, and the return type of fluent setters in
\texttt{Database} is automatically refined to \texttt{ExtendedDatabase}:
\lstinputlisting[linerange=14-18]{../UseMixinLombok/src/casestudy/sql/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE_EXT
In this way, when a query is created using \texttt{ExtendedDatabase},
all the fluent setters return the correct type, and not the old \texttt{Database} type, which would prevent calling
\texttt{orderBy}.

@@

\subsection{A Maze Game}
This case study is a simplified variant of a Maze game, which is often
used~\cite{gof,bono14}
to evaluate code reuse ability related to inheritance and design
patterns. In the game, there is a player with the goal of collecting
as many coins as possible. She may enter a room with several doors to
be chosen among. This is a good example because it involves code reuse
(different kinds of doors inherit a common type, with different
features and behavior), multiple inheritance (a special kind of door
may require features from two other door types) and it also shows how
to model operations \texttt{symmetric sum}, \texttt{override} and
\texttt{alias} from trait-oriented programming. The game has been
implemented using plain Java 8 and default methods by Bono
et. al~\cite{bono14}, and the code for that implementation is
available online. We reimplemented the game using \mixin. Due to space
constraints, we omit the code here. The following table summarizes
the number of lines of code and classes/interfaces in each implementation:

>>>>

As the first application for Classless Java, we illustrate a useful
programming pattern that improves modularity and extensibility of
programs. This useful pattern is based on an existing solution to
the \emph{Expression Problem} (EP)~\cite{wadler98expression}, which is a well-known
problem about modular extensibility issues in software evolution. Recently, a
new solution~\cite{eptrivially} using only covariant type refinement was
proposed. When this solution is modelled with interfaces and default methods, it
can even provide independent extensibility~\cite{zenger05independentlyextensible}: the ability to assemble a system
from multiple, independently developed extensions. Unfortunately, the
required instantiation code makes a plain Java solution verbose and cumbersome
to use. The \mixin annotation is enough to remove the boilerplate code, making
the presented approach very appealing. Our last case study, presented
in Section~\ref{subsec:int}, is essentially a (much larger) application of this
pattern to an existing program. Here we illustrate the pattern in the
much smaller Expression Problem.

\paragraph{Initial System}
In the formulation of the EP, there is an initial system that models
arithmetic expressions with only literals, addition, and an initial
operation \texttt{eval} for expression evaluation.
As shown in Figure~\ref{fig:ep}, \texttt{Exp} is the common
super-interface with operation \texttt{eval()}
inside. Sub-interfaces \texttt{Lit} and \texttt{Add} extend interface
\texttt{Exp} with default implementations for the \texttt{eval} operation. The
number field \texttt{x} of a literal is represented as a getter method
\texttt{x()} and expression fields (\texttt{e1} and \texttt{e2}) of an addition
as getter methods \texttt{e1()} and \texttt{e2()}.

@@

\paragraph{Extending the Query Language} The previous query language can be extended with a new feature
\texttt{orderBy} which orders the result records by a field that users
specify. With \mixin programmers just need to extend the interface \texttt{Database} with new
features, and the return type of fluent setters in
\texttt{Database} is automatically refined to \texttt{ExtendedDatabase}:
\lstinputlisting[linerange=14-18]{../UseMixinLombok/src/casestudy/sql/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE_EXT
In this way, when a query is created using \texttt{ExtendedDatabase},
all the fluent setters return the correct type instead of the old \texttt{Database} type, which would prevent calling
\texttt{orderBy}.

@@

\subsection{A Maze Game}
This case study is a simplified variant of a Maze game, which is often
used~\cite{gof,bono14}
to evaluate code reuse ability related to inheritance and design
patterns. In the game, there is a player with the goal of collecting
as many coins as possible. She may enter a room with several doors to
be chosen among. This is a good example because it involves code reuse
(different kinds of doors inherit a common type, with different
features and behavior), multiple inheritance (a special kind of door
may require features from two other door types) and it also shows how
to model operations \texttt{symmetric sum}, \texttt{override} and
\texttt{alias} from trait-oriented programming. The game has been
implemented using plain Java 8 and default methods by Bono
et. al~\cite{bono14}, and the code for that implementation is
available online. We refactored the game using \mixin. Due to space
constraints, we omit the code here. The following table summarizes
the number of lines of code and classes/interfaces in each implementation:

### 2016-06-30 GPCE2016/sections/Implementation.tex
grammar

Disciplined use of Classless Java (avoiding class
declarations as done in Section~\ref{sec:ep}) illustrates what \emph{pure} IB is like.
However, using \mixin, CB and IB programming can be mixed together,
harvesting the practical convenience of using existing Java libraries, the full
Java language and IDE support.
The key to our implementation are compilation agents, which
 allow us to rewrite the Java AST just
before compilation. We discuss advantages and limitations of our approach.

@@

\paragraph{Modularity}
While general preprocessing acts across module boundaries, compilation
agents act modularly on each class/compilation unit. It makes sense to
apply the transformations to one class/interface at a time, and only to
annotated classes/interfaces. This allows library code to be reused
without being reprocessed and recompiled, making our
approach 100\% compatible with existing Java libraries, which can be
used and extended normally.

@@

\item{\bf Self coherence}: the generated code itself is well-typed. That is,
  type errors are not present in code the user has not written (for
  example \texttt{of} methods in Figure~\ref{fig:screenshot}).
In our case, it means that either \mixin{} produces (in a controlled way) an
understandable error or the interface can be successfully annotated and the generated code is well-typed.

@@

\subsection{Limitations}
Our prototype implementation has certain limitations:
\begin{itemize}
\item Lombok allows writing handlers for either javac or ejc(Eclipse's own compiler).
Our current implementation only realizes ejc version. The implementation for
  the \texttt{javac} version is still missing.
\item Simple generics is supported:
type parameters can be used, but generic
method typing is %neither formalized in this paper nor 
not explicitly checked by \mixin, but simply delegated to the Java compiler.
\item 
Due to limited support in Lombok for separate compilation, i.e., 
accessing information of code defined in different files, \mixin 
%In the same spirit, we have a mature \mixin annotation, which
%does not support separate compilation yet: it 
requires that all
  related interfaces have to appear in a single Java file.
Reusing the logic inside the experimental Lombok annotation \lstinline{@Delegate},
we also offer a less polished annotation supporting
separate compilation.
%  has the same behavior as \mixin. A limitation is
%  that users need to put references to the super types
% of the annotated type together with the annotation, for instance,
%  \lstinline{@Delegate(types = Point2D.class)}.
%   Another limitation is the methods generated by \lstinline{@Delegate} will
%  not be visible in the Outline window, but they can still be auto-completed in Eclipse.
\end{itemize}

>>>>

Disciplined use of Classless Java (avoiding class
declarations as done in Section~\ref{sec:ep}) illustrates what \emph{pure} IB is like.
However, using \mixin, CB and IB programming can be mixed together,
harvesting the practical convenience of using existing Java libraries, the full
Java language and IDE support.
The key to our implementation is compilation agents, which
 allow us to rewrite the Java AST just
before compilation. We discuss the advantages and limitations of our approach.

@@

\paragraph{Modularity}
While general preprocessing acts across module boundaries, compilation
agents act modularly on each class/compilation unit. It makes sense to
apply the transformations to one class/interface at a time, and only to
annotated classes/interfaces. This allows library code to be reused
without being reprocessed or recompiled, making our
approach 100\% compatible with existing Java libraries, which can be
used and extended normally.

@@

\item{\bf Self coherence}: the generated code itself is well-typed. 
In our case, it means that either \mixin{} produces (in a controlled way) an
understandable error or the interface can be successfully annotated and the generated code
 (e.g. the \texttt{of} methods in Figure~\ref{fig:screenshot}) is well-typed.

@@

\subsection{Limitations}
Our prototype implementation has certain limitations:
\begin{itemize}
\item Lombok allows writing handlers for either javac or ejc(Eclipse's own compiler).
Our current implementation only realizes ejc version. The implementation for
  the \texttt{javac} version is still missing.
\item Simple generics is supported:
type parameters can be used, but generic method typing is 
delegated to the Java compiler instead of  %neither formalized in this paper nor 
being explicitly checked by \mixin.
\item 
Due to limited support in Lombok for separate compilation, i.e., 
accessing information of code defined in different files, \mixin 
%In the same spirit, we have a mature \mixin annotation, which
%does not support separate compilation yet: it 
requires that all
  related interfaces have to appear in a single Java file.
Reusing the logic inside the experimental Lombok annotation \lstinline{@Delegate},
we also offer a less polished annotation supporting
separate compilation.
%  has the same behavior as \mixin. A limitation is
%  that users need to put references to the super types
% of the annotated type together with the annotation, for instance,
%  \lstinline{@Delegate(types = Point2D.class)}.
%   Another limitation is the methods generated by \lstinline{@Delegate} will
%  not be visible in the Outline window, but they can still be auto-completed in Eclipse.
\end{itemize}

### 2016-06-30 GPCE2016/sections/Overview.tex
grammar

\noindent Only code for methods needing implementation is generated. Thus,
programmers can easily customize the behaviour for their special needs.
Also, since \mixin interfaces offer the \Q@of@ factory method, only interfaces where all the abstract methods
can be synthesized can be object interfaces. A non \mixin interface is like an abstract class in Java.
%\end{comment}

@@

\begin{figure*}
\saveSpaceFig
\centering
{%\fontsize{8}{10}\selectfont
\begin{tabular}{|l|l|l|l|}
\hline
& \textbf{Operation}  & \textbf{Example}                  & \textbf{Description } \\ \hline
\multirow{4}{*}{\parbox{2.5cm}{State operations (for a field \texttt{x})}} & \textbf{``fields''/getters}        &   \Q@int x()@                  & Retrieves value from field \texttt{x}.          \\ \cline{2-4}
& {\bf withers}        &   \Q@Point2D withX(int val)@                & Clones
object; updates field \texttt{x} to \texttt{val}.             \\ \cline{2-4}
& \textbf{setters}        & \Q@void x(int val)@ & Sets the field
\texttt{x} to a  new value \texttt{val}.        \\ \cline{2-4}
& \textbf{fluent setters}        & \Q@Point2D x(int val)@ &Sets the field
\texttt{x} to \texttt{val} and returns \texttt{\this}.           \\ \hline
\multirow{3}{*}{Other operations} &
\textbf{factory methods} &
%\begin{tabular}{l}
\Q@static Point2D of(int _x,int _y)@
%\end{tabular}
 & Factory method (generated).        \\
\cline{2-4}
& \textbf{functional updaters}        & \Q@Point3D with(Point2D val)@
& Update all matching fields in \texttt{val}.        \\ \hline
%\cline{2-4}
%& \textbf{imperative updaters}        & \Q@Point3D set(Point2D val)@ &
%Set all matching fields in \texttt{val}.        \\ \hline
\end{tabular}
}
\caption{Abstract state operations for a field \texttt{x}, together with other operations, supported by the \mixin
  annotation. }

>>>>

\noindent Only code for methods needing implementation is generated. Thus,
programmers can easily customize the behaviour for their special needs.
Also, since \mixin interfaces offer the \Q@of@ factory method, only interfaces where all the abstract methods
can be synthesized can be object interfaces. A non-\mixin interface is like an abstract class in Java.
%\end{comment}

@@

\begin{figure*}
\saveSpaceFig
\centering
{%\fontsize{8}{10}\selectfont
\begin{tabular}{|l|l|l|l|}
\hline
& \textbf{Operation}  & \textbf{Example}                  & \textbf{Description } \\ \hline
\multirow{4}{*}{\parbox{2.5cm}{State operations (for a field \texttt{x})}} & \textbf{``fields''/getters}        &   \Q@int x()@                  & Retrieves value from field \texttt{x}.          \\ \cline{2-4}
& {\bf withers}        &   \Q@Point2D withX(int val)@                & Clones
object; updates field \texttt{x} to \texttt{val}.             \\ \cline{2-4}
& \textbf{setters}        & \Q@void x(int val)@ & Sets the field
\texttt{x} to a  new value \texttt{val}.        \\ \cline{2-4}
& \textbf{fluent setters}        & \Q@Point2D x(int val)@ &Sets the field
\texttt{x} to \texttt{val} and returns \texttt{\this}.           \\ \hline
\multirow{3}{*}{Other operations} &
\textbf{factory methods} &
%\begin{tabular}{l}
\Q@static Point2D of(int _x,int _y)@
%\end{tabular}
 & Factory method (generated).        \\
\cline{2-4}
& \textbf{functional updaters}        & \Q@Point3D with(Point2D val)@
& Updates all matching fields in \texttt{val}.        \\ \hline
%\cline{2-4}
%& \textbf{imperative updaters}        & \Q@Point3D set(Point2D val)@ &
%Set all matching fields in \texttt{val}.        \\ \hline
\end{tabular}
}
\caption{Abstract state operations for a field \texttt{x}, together with other operations, supported by the \mixin
  annotation. }

### 2016-06-30 GPCE2016/sections/RelatedWork.tex
grammar

%\paragraph{Mixins.}
Mixins are more restricted than the C++ approach. Mixins allow naming
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearization: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearization (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits~\cite{scala-overview} are in fact more like linearized mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as the diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearized mixin composition. % Other languages, such as
% Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearized mixins. Indeed, in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@
\end{comment}
\noindent Java interfaces and default methods do not use
linearization: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

>>>>

%\paragraph{Mixins.}
Mixins are more restricted than the C++ approach. Mixins allow naming
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearization: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearization (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight against this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits~\cite{scala-overview} are in fact more like linearized mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as the diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearized mixin composition. % Other languages, such as
% Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearized mixins. Indeed, in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@
\end{comment}
\noindent Java interfaces and default methods do not use
linearization: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.
